---
title: "Mini Project 3"
author: "Benny Frisella"
date: "2024-10-20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(root.dir = "C:/Users/benny/Documents/University of Texas @ Dallas/Statistical Learning/Datasets")
```

##1. Consider the diabetes dataset from Mini Project 2. Use all predictors for all the models considered
##for this problem.

(a) Fit a logistic regression model using all predictors in the data. Provide its error rate, sensitivity,
and specificity based on training data.
```{r}
#(a)

setwd("~/University of Texas @ Dallas/Statistical Learning/Datasets")
diabetes <- read.csv("diabetes.csv")
# summary(diabetes) -> medium-sized set
# colnames(diabetes)
logReg <- glm(Outcome ~ Pregnancies.. + Glucose.. + BloodPressure.. + SkinThickness.. + Insulin.. + BMI.. + DiabetesPedigreeFunction.. + Age.., family = binomial, data = diabetes)

#Randomly split data into: 70% training and 30% prediction sets
#set.seed(8)
sampleSize <- floor(0.7 * nrow(diabetes))
trainIndex <- sample(seq(nrow(diabetes)), size = sampleSize)

#include randomly chosen indices for training
trainData <- diabetes[trainIndex,]

#include what wasn't chosen in training set into test set
testData <- diabetes[-trainIndex,]



#Build model on training set
fit <- glm(Outcome ~ Pregnancies.. + Glucose.. + BloodPressure.. 
           + SkinThickness.. + Insulin.. + BMI.. 
           + DiabetesPedigreeFunction.. + Age.., 
           family = binomial, data = trainData)

summary(fit)

#estimated probabilities for test data
logReg.prob <- predict(fit, testData, type = "response")

#predicted classes (using 0.5 cutoff)
logReg.pred <- ifelse(logReg.prob >= 0.5, 1, 0)

#test error rate
1 - mean(logReg.pred == testData[, "Outcome"])



# Confusion matrix and (sensitivity, specificity)
# `+' = 1, `-' = 0

table(logReg.pred, testData[, "Outcome"])

errorRate <- (37 + 92) / (353 + 118 + 37 + 92)
sensitivity <- 118 / (118 + 92)
specificity <- 353 / (353 + 37)

cat("\nError rate: (FP + FN) / (TP + TN + FP + FN) =", errorRate)
cat("\nSensitivity: TP / (TP + FN)", sensitivity)
cat("\nSpecificity: TN / (TN + FP)", specificity)

```
(b) Write your own code to estimate the test error rate of the model in (a) using LOOCV

```{r}
#(b)

#observations
n <- nrow(diabetes)

squaredErrorsLoocv <- numeric(n)

#LOOCV
for (i in 1:n) 
{
  #create training/test sets
  trainData <- diabetes[-i, ]
  testData <- diabetes[i, , drop = FALSE]
  
  #train using logistic regression
  LOOCV <- glm(Outcome ~ Pregnancies.. + Glucose.. + BloodPressure.. 
                     + SkinThickness.. + Insulin.. + BMI.. 
                     + DiabetesPedigreeFunction.. + Age..,
                     data = train_data, family = binomial)
  
  #make predictions for the i-th test case
  prediction <- predict(LOOCV, newdata = testData, type = "response")
  
  #square the difference, store
  actualOutcomes <- testData$Outcome
  squaredErrorsLoocv[i] <- (prediction - actualOutcomes) ^ 2
}

#Mean Squared Error for LOOCV
LOOCV_MSE <- mean(squaredErrorsLoocv)

LOOCV_MSE

```

```{r}
#(c)

library(boot)

#Build glm model on entire dataset
model <- glm(Outcome ~ Pregnancies.. + Glucose.. 
             + BloodPressure.. + SkinThickness.. + Insulin.. 
             + BMI.. + DiabetesPedigreeFunction.. + Age.., 
             family = binomial, data = diabetes)

#k-fold LOOCV special case
KFoldsLOOCV <- cv.glm(diabetes, model, K = n)

KFoldsLOOCV

```

```{r}
?cv.glm
```

```{r}
# k = 10 folds
KFolds10 <- cv.glm(diabetes, model, K = 10)

KFolds10
```
2. Consider the oxygen saturation data stored in oxygen_saturation.txt file. The data consist of
measurements of percent saturation of hemoglobin with oxygen in 72 adults, obtained using an oxygen saturation monitor (OSM, method 1) and a pulse oximetry screener (POS, method 2). You can read
about oxygen saturation on Wikipedia, https://en.wikipedia.org/wiki/Oxygen_saturation_
(medicine). We are primarily interested in evaluating agreement between the two methods for
measuring oxygen saturation.

(a) Let Y1 and Y2 denote the population of observations of methods 1 and 2, respectively, and
D = Y1 − Y2 denote their difference. Let theta be the total deviation index (TDI) between the two
methods. For a given large probability p, it is defined as the pth quantile of |D|. Here we will
take p = 0.90. Argue that smaller values for theta imply better agreement.
```{r}
#(a)

setwd("~/University of Texas @ Dallas/Statistical Learning/Datasets")
oxygenSats <- read.csv("oxygenSaturation.txt", sep = "\t", header = TRUE)

oxygenSats$D <- abs(oxygenSats$pos - oxygenSats$osm)

hist(oxygenSats$D, main = "Total Deviation Index Distribution", xlab = "Differences Between POS and OSM")
```
(b) Provide a point estimate theta_hat of theta. (Tip: If the population parameter is a quantile, what should be
its natural estimator?) Write your own code to compute (nonparametric) bootstrap estimates
of bias and standard error of theta_hat, and a 95% upper confidence bound for theta computed using the
percentile method. Interpret the results.
```{r}
#(b)

#set.seed(8)

#point-estimation
thetaHat <- quantile(oxygenSats$D, 0.9)

#   Bootstrap NonParametric   #
B <- 1000  #number of samples
bootstrapEstimates <- numeric(B)

for (b in 1:B) 
{
  bootstrapSample <- sample(oxygenSats$D, replace = TRUE)
  bootstrapEstimates[b] <- quantile(bootstrapSample, 0.9)
}

#bias
bias <- mean(bootstrapEstimates) - thetaHat

#standard error
sdError <- sd(bootstrapEstimates)

# 95% upper confidence bound
upperBound <- quantile(bootstrapEstimates, 0.95)

cat("\nPoint estimate (theta_hat):", thetaHat, "\n")
cat("Bias:", bias, "\n")
cat("Standard Error:", sdError, "\n")
cat("95% Upper Confidence Bound:", upperBound, "\n")
```
(c) Repeat the computation in (d) using boot package in R, or check bootstrap function from the
SciPy library ( scipy.stats.bootstrap) in Python, and compare your results.
```{r}
#(c)
library(boot)

statistic <- function(data, indices) 
{
  quantile(abs(data[indices, 1] - data[indices, 2]), 0.9)
}

bootstrap <- boot(oxygenSats, statistic, R = 1000)

bootstrap
```

