---
title: "Mini Project 6"
author: "Benny Frisella"
date: "2024-12-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
**1. Consider the Hitters dataset from the previous project. It consists of 20 variables measured on
263 major league baseball players (after removing those with missing data). Take log(Salary) as
response (due to skewness in Salary) and the remaining 19 variables as predictors. All data will
be taken as training data. For all the models below, use leave-one-out cross-validation (LOOCV) to
compute the estimated test MSE.


(a) Fit a tree to the data. Summarize the results. Unless the number of terminal nodes is large,
display the tree graphically and explicitly describe the regions corresponding to the terminal
nodes that provide a partition of the predictor space (i.e., provide expressions for the regions
R1, . . . , RJ ). Report its estimated test MSE.

```{r}
library(ISLR)
library(tree)

#?tree

##       Question 1       ##

# (a)

#omit NA
Hitters <- na.omit(Hitters)

summary(Hitters)

#convert categorical columns to factor
categorical <- which(sapply(Hitters, is.character))

Hitters[categorical] <- lapply(Hitters[categorical], as.factor)

Hitters.formula <- log(Salary) ~ .

#str(Hitters)

#create log() Regression Tree #
tree.Hitters <- tree(log(Hitters$Salary) ~ ., Hitters)

summary(tree.Hitters)

#plot
plot(tree.Hitters)
text(tree.Hitters, pretty = 0)

#use LOOCV for test MSE
LOOCV.tree <- cv.tree(tree.Hitters, K = nrow(Hitters))
tree.MSE <- mean(LOOCV.tree$dev / length(Hitters$Salary))  #average MSE

tree.MSE


```

```{r}

# (b)

#try cost complexity pruning by LOOCV
which.min(LOOCV.tree$size)

#plot estimated test error
plot(LOOCV.tree$size, LOOCV.tree$dev, type = "b", main = "Cost Complexity Pruning",
     ylab = "test error", xlab = "Size")
     mtext("(Using LOOCV)", side = 4, line = 1, cex = 1)

prune.salary <- prune.tree(tree.Hitters, best = 8)

plot(prune.salary)
text(prune.salary, pretty = 0)

#use LOOCV for test MSE
LOOCV.tree <- cv.tree(prune.salary, K = nrow(Hitters))
prune.MSE <- mean(LOOCV.tree$dev / length(Hitters$Salary))  #average MSE

prune.MSE

```

```{r}

# (c)

library(randomForest)
library(caret)

#bagging model w/ B = 1000 trees
bagging.model <- randomForest(log(Salary) ~ ., data = Hitters, ntree = 1000, mtry = ncol(Hitters) - 1)

bagging.model

#get test MSE
testMSE <- bagging.model$mse[1000]  #get value for the last tree
cat("Bagging Test MSE:", testMSE, "\n")

#get important predictors
important.values <- importance(bagging.model)
print(important.values)

#plot
varImpPlot(bagging.model, main = "Bagging")

#Bagging with LOOCV
bag_model <- train(log(Salary) ~ ., data = Hitters, method = "treebag", trControl = ctrl)

print(bag_model) #rquared = 71%

#get predictions
predictions_bag <- bag_model$pred$pred

#get actual values
actual_bag <- bag_model$pred$obs

#get MSE
mse_bag <- mean((predictions_bag - actual_bag)^2)
print(mse_bag)

```

```{r}

# (d)

#random forest with m ≈ p/3 (default)
randForest.model <- randomForest(log(Salary) ~ ., data = Hitters, ntree = 1000)

# Check Random Forest results
print(randForest.model)

# Estimated Test MSE (Out-of-Bag Error)
randForest.model$mse[1000]

#plot
varImpPlot(randForest.model, main = "Random Forest")



#prep
ctrl <- trainControl(method = "LOOCV")

#Random Forest with LOOCV
rf_model <- train(log(Salary) ~ ., data = Hitters, method = "rf", trControl = ctrl)

#get predictions
predictions <- rf_model$pred$pred

#get actual values
actual <- rf_model$pred$obs

#get MSE
mse_rand <- mean((predictions - actual)^2)
print(mse_rand)



```


```{r}
#combine
layout(matrix(c(1, 2), nrow = 1, ncol = 2))

varImpPlot(bagging.model, main = "Bagging")
varImpPlot(randForest.model, main = "Random Forest")

```

```{r}

# (e)

library(gbm)

#fit boosting with LOOCV
boosting.model <- train(log(Salary) ~ ., data = Hitters, 
                        method = "gbm", trControl = trainControl("LOOCV"),
                        tuneGrid = expand.grid(interaction.depth = 1,
                                               n.trees = 1000, 
                                               shrinkage = 0.01, #lambda
                                               n.minobsinnode = 10),
                        verbose = FALSE)

#relative influence
summary(boosting.model)

#see
boosting.model$results

#get MSE
boostMSE <- boosting.model$results$RMSE^2  # RMSE squared gives MSE
print(boostMSE)


```

```{r}

# (f)

#create test MSE table
model_comparison <- data.frame(
  Model = c("Random Forest", "Boosting", "Bagging", "Un-Pruned Tree", "Pruned Tree"),
  TestMSE = c(mse_rand, boostMSE, mse_bag, tree.MSE, prune.MSE)
)

# Print the comparison table
print(model_comparison)

```

```{r}
### Question 2 ###

# (a)
setwd("~/University of Texas @ Dallas/Statistical Learning/Datasets")

#prep
diabetes <- read.csv("diabetes.csv")

#factor needed
diabetes$Outcome <- as.factor(diabetes$Outcome)

#level error fix
levels(diabetes$Outcome) <- c("Class0", "Class1")

formula <- Outcome ~ .

#10-fold control
ctrl <- trainControl(method = "cv", number = 10, classProbs = TRUE, summaryFunction = twoClassSummary)

#fit SVC
svc_model <- train(formula, data = diabetes, method = "svmLinear", trControl = ctrl, metric = "ROC",
                   tuneLength = 10)
svc_model

#test error
svc_pred <- predict(svc_model, diabetes)
svc_error <- mean(svc_pred != diabetes$Outcome)
cat("SVC Test Error Rate:", svc_error, "\n")

#tune cost

#grid of cost values for tuning
cost_grid <- expand.grid(C = seq(0.001, 0.5, by = 0.005)) #CHANGE THIS, BEST = 0.011

svc_model <- train(formula, data = diabetes, method = "svmLinear", trControl = ctrl, 
  tuneGrid = cost_grid,
  metric = "ROC")

#find best model
print(svc_model)

#get cost
best_cost <- svc_model$bestTune$C
cat("Best cost parameter:", best_cost, "\n")

#use cost
tuned_grid <- expand.grid(C = 6.501)

#train again
svm_model <- train(Outcome ~ ., data = diabetes, method = "svmLinear", trControl = ctrl, 
                   tuneGrid = tuned_grid)

print(svm_model)

```

```{r}
# (b)

#SVM
set.seed(123)
poly_svm_model <- train(Outcome ~ ., data = diabetes, method = "svmPoly",trControl = ctrl,
  tuneGrid = expand.grid(degree = 2, scale = seq(0.1, 1, length = 5), C = seq(0.01, 1, length = 5)),
  metric = "ROC")

#show
print(poly_svm_model)
poly_svm_model$bestTune  #optimal
poly_svm_model$results   #performance

#test MSE
svm_pred <- predict(poly_svm_model, diabetes)
svm_error <- mean(svm_pred != diabetes$Outcome)
cat("SVM Test Error Rate:", svm_error, "\n")


```

```{r}
# (c)

#radial SVM
set.seed(123)
rbf_svm_model <- train(Outcome ~ ., data = diabetes, method = "svmRadial", trControl = ctrl,
  tuneGrid = expand.grid(C = seq(0.01, 1, length = 5), sigma = seq(0.01, 0.1, length = 5)),
  metric = "ROC")

#get results
print(rbf_svm_model)
rbf_svm_model$bestTune  #optimal parameters
rbf_svm_model$results   #performance

#test MSE
rbf_pred <- predict(rbf_svm_model, diabetes)
rbf_error <- mean(rbf_pred != diabetes$Outcome)
cat("RBF Test Error Rate:", rbf_error, "\n")


```

```{r}

# (e) compare
roc_linear <- 0.83078
sens_linear <- 0.89356
spec_linear <- 0.54539

roc_poly <- 0.85817
sens_poly <- 0.90880
spec_poly <- 0.58618

roc_radial <- 0.88542
sens_radial <- 0.90729
spec_radial <- 0.65650

sv_results <- data.frame(
  Model = c("SVC", "SVM(Degree = 2)", "SVM(Radial)"),
  ROC = c(roc_linear, roc_poly, roc_radial),
  Sensitivity = c(sens_linear, sens_poly, sens_radial),
  Specificity = c(spec_linear, spec_poly, spec_radial),
  Misclassifcation_Rate = c(svc_error, svm_error, rbf_error)
)

# Print the results table
print(sv_results)

```

```{r}

### Question 3 ###

#(c)

library(factoextra) # For visualization

#omit NA
Hitters <- na.omit(Hitters)

#only work with numeric
Hitters_numeric <- Hitters[, sapply(Hitters, is.numeric)]

#standardize
data_standardized <- scale(Hitters_numeric)

#Hierarchical clustering w/ Euclidean distance/complete linkage
diss_matrix <- dist(data_standardized, method = "euclidean")
hclust_model <- hclust(diss_matrix, method = "complete")

#cut model create 2 clusters
clusters <- cutree(hclust_model, k = 2)

#get means
cluster_summary <- aggregate(Hitters[, -c(1)], by = list(Hitters$Cluster), FUN = mean)

salary_summary <- aggregate(Hitters$Salary, by = list(Hitters$Cluster), FUN = mean)

print(cluster_summary)
print(salary_summary)

#dress down/up tree
library(dendextend)

#create dendogram
dend <- as.dendrogram(hclust_model)

#remove leafs
dend <- dend %>% set("labels", rep("", length(labels(dend))))

#plot
plot(dend, 
     main = "Hierarchical Clustering", 
     xlab = "", 
     ylab = "Height", 
     sub = "",
     cex = 0.8)
```

```{r}

# (d)


#K-means clustering w/ K = 2
set.seed(123)  # For reproducibility
kmeans_model <- kmeans(data_standardized, centers = 2)

#add labels
Hitters$Cluster_Kmeans <- as.factor(kmeans_model$cluster)

#summarize means
kmeans_summary <- aggregate(Hitters[, -c(1, 18)], by = list(Hitters$Cluster_Kmeans), FUN = mean)

#summarize mean salaries
salary_summary_kmeans <- aggregate(Hitters$Salary, by = list(Hitters$Cluster_Kmeans), FUN = mean)

print(kmeans_summary)
print(salary_summary_kmeans)


```